---
title: "Column Selection"
format: html
editor: visual
---

**PURPOSE:**  The purpose of this report is to examine and select columns from COMBINED_DEGREE_V table for further use and modeling.  It examines columns for similarity (such as "MAJOR" and "MAJOR_CD") and variance. 

# Preliminary

Columns were initially selected from a summary report of the entire database.  They were selected based on having complete or nearly complete data (a low percentage of missing values.)


```{r}

##########  
## LOAD ##  
########## 

gradData <- readRDS(here::here("Data", "gradData_early_pull.rds"))

```

```{r}

###############
## FUNCTIONS ##
###############

library(kableExtra)

detSim <- function(col1,col2) {

  # This function "determines similarity" between columns to determine if the columns are redundant or in other words if they are different labels for the same thing.  For example, "COLLEGE" and "COLLEGE_SHORT" should be completely redundant.  However, "COLLEGE" and "ACAD_COLLEGE_CIP_CD" might not be redundant.  
    
  testTable <- table(col1,col2) |>
    as.data.frame.matrix()
  
  # Every row and every column has exactly one non-zero cell
  identicalTest <- 
    all(apply(testTable, 1, function(x) sum(x > 0) == 1)) &&
    all(apply(testTable, 2, function(x) sum(x > 0) == 1))
  
  if(identicalTest) {return(
    list(sim = "Identical", row = NA, col = NA))
    }
  
  # return row and columns that are off 
  
  offRow <- apply(testTable, 1, function(x) sum(x > 0)) |> (\(x){which(x>1)})()
  offCol <- apply(testTable, 2, function(x) sum(x > 0)) |> (\(x){which(x>1)})()
  
  # percentage difference
  row_max_sum <- sum(apply(testTable, 1, max))
  col_max_sum <- sum(apply(testTable, 2, max))
  row_sim <- row_max_sum / sum(testTable)
  col_sim <- col_max_sum / sum(testTable)
  
  return(list(sim = min(row_sim, col_sim), row = offRow, col = offCol))
}


createCombinatorics <- function(vec) {

  # This function receives a vector of column names and then produces a grid of every dual combination.  It will be used to calculate the similarity
    
  full_grid <- expand.grid(vec,vec,stringsAsFactors = FALSE)
  reduced_grid <- full_grid[full_grid$Var1 < full_grid$Var2,]
  
  return(reduced_grid)
  
}


family_detSim <- function(fam, data){
  
  # where fam is a vector of column names chosen for their suspected similarity
  # where data is the dataframe where the columns occur
  
sim_list <- fam |>
createCombinatorics() |>
  (\(comb){
    setNames(
  lapply(1:nrow(comb),
         function(y){
           detSim(data[, comb[y,"Var1"]], data[,comb[y,"Var2"]])
         }
         ),
  paste(comb[,"Var1"],comb[,"Var2"])
    )
  })()

  return(sim_list)

}

## MODIFIED CLEANING FUNCTION ##  

cleanDiff <- function(majr, majr_short) {
  
  short_intermediate <- majr_short |>
    (\(x)({gsub("Psychology Teaching", "Psychology", x)}))() |>
    (\(x)({gsub("Health Promotion & Education Teaching", "Health Promotion & Education", x)}))() |>
  (\(x)({gsub("Consumer & Community Studies", "Consumer & Communities Studies", x)}))() |>
  (\(x)({gsub("Parks, Recreation & Tourism", "Parks, Recreation, & Tourism", x)}))() |>
    (\(x)({gsub("Health, Society & Policy", "Health, Society, & Policy", x)}))()
  
  charDiff <- mapply(function(x,y) {gsub(x, "", y) }, x = short_intermediate, y = majr)
  
  # return(short_intermediate)
  
  return(charDiff)
  
}


```


```{r}

#####################
## COLUMN FAMILIES ##
#####################

# columns that sound like they are similar

div_fam <- c(
  "ACAD_DIVISION_CD",
  "DIVISION",
  "DIVISION_SHORT", 
  "DIVISION_FORMAL",
  "ACAD_DIVISION_CIP_CD",
  "DIVISION_CD"
)


college_fam <- c(
  "COLLEGE_SHORT",
  "ACAD_COLLEGE_CIP_CD", 
  "COLLEGE_CD",
  "SNAP_MAJORCOLL"
)

dept_fam <- c(
   "MAJOR_ACAD_ORG_CD",
   "ACAD_DEPARTMENT_CD",
    "DEPARTMENT", 
   "DEPARTMENT_SHORT",
   "DEPARTMENT_FORMAL",
  "ACAD_DEPARTMENT_CIP_CD",
  "DEPARTMENT_CD", 
  "SNAP_MAJORDEPT"
)

major_fam <- c("MAJOR_CD", 
               "MAJOR",
               "MAJOR_SHORT", 
               "MAJOR_LONG",
               "MAJOR_GROUP",
               "CIP_CD",
               "MAJOR_ACAD_ORG_CD",
               "SNAP_MAJORCDE",
               "SNAP_MAJORNAME",
               "SNAP_CIP"
               )

```


```{r}

# college_sim <- family_detSim(fam=college_fam, data = gradData)

simList <- lapply(list(
  div=div_fam,
  college =college_fam,
  dept = dept_fam,
  major = major_fam
            ),
       family_detSim, data = gradData)


```

# Results

## Columns per 'Family' Selected

### **Divisions:**  

The columns "ACAD_DIVISION_CD",  "DIVISION_CD", "DIVISION", "DIVISION_SHORT", and "DIVISION_FORMAL" are identical.  They are 98% similar to "ACAD_DIVISION_CIP_CD".

For clarity and ease of interpretation, I will select "DIVISION_SHORT" and abandon the rest.


### **Colleges:**

The columns "ACAD_COLLEGE_CIP_CD", "COLLEGE_SHORT", and "COLLEGE_CD" are identical.  They are 98% similar to "SNAP_MAJORCOLL."

For clarity and ease of interpretation, I will select "COLLEGE_SHORT" and abandon the other three.  

### **Departments:**  

The columns "DEPARTMENT", "DEPARTMENT_SHORT", "DEPARTMENT_FORMAL",  "ACAD_DEPARTMENT_CD", and "DEPARTMENT_CD" are identical.

These columns are 98.6% similar to ACAD_DEPARTMENT_CIP_CD

These columns are 97.6% similar to MAJOR_ACAD_ORG_CD

These columns are 87.2% similar to SNAP_MAJORDEPT.

ACAD_DEPARTMENT_CIP_CD is 96.3% similar to MAJOR_ACAD_ORG_CD and 86.4% similar to SNAP_MAJORDEPT.

MAJOR_ACAD_ORG_CD is 88.1% similar to SNAP_MAJORDEPT.

For clarity and ease of interpretation, I will select "ACAD_DEPARTMENT_CD" and abandon the rest.  

### **Majors:**  

Out of the `r length(major_fam)` columns related to major, only two are identical:  MAJOR_CD and SNAP_MAJORCDE.

The remaining columns vary between 47% similarity (MAJOR_ACAD_ORG_CD and SNAP_MAJORNAME) and 99.8% (MAJOR_GROUP and MAJOR_SHORT).  Each column's mean similarity to the other columns is show in  [FIGURE] below. 

The column "MAJOR_SHORT" has the second highest mean similarity to the other columns, and will be selected for brevity and clarity.  It differs from "MAJOR", "MAJOR_LONG", and "MAJOR_CD" in that these columns add details about the degree awarded.  With some manipulation, these details are added as two additional columns: "award" and "honors."  

The column "MAJOR_CD" may be convenient for labels on plots.  

For clarity and ease of interpretation, I will select "MAJOR_SHORT", "MAJOR" (to extract the degree awarded), and "MAJOR_CD" (for plot labels) and abandon the rest.  

```{r warning = FALSE}

## MAJOR ## 

majSim <- lapply(simList[["major"]], function(x) { x[["sim"]]}) |> unlist() |> (\(x){x[order(x, decreasing = TRUE)]})()


mean_val <- lapply(major_fam, function(maj){
  
  majSim[grepl(maj, names(majSim))] |>
    as.numeric() |>
    mean(na.rm=TRUE)
}
)
names(mean_val) <- major_fam

mean_val <- mean_val |> 
  unlist() |>
  (\(x){x[order(x, decreasing = TRUE)]})() 

```


```{r include=FALSE}

plot(mean_val,
       las = 1,
       ylab = "Mean Similarity",
     xaxt = "n",
     xlab = "",
     main = "Mean similarity between 'major' family columns"
       )
grid()

text(x = 1:(length(mean_val)-2),
     y = mean_val[-(length(mean_val)-1):-length(mean_val) ] - 0.02,
     labels = names(mean_val)[-(length(mean_val)-1):-length(mean_val) ],
     srt = 90, cex = 0.6, font = 2,
     adj = c(1, 0.5))

text(x = length(mean_val)-1.1,
     y = mean_val[length(mean_val)-1],
     labels = names(mean_val)[length(mean_val)-1],
     srt = 90, 
     cex = 0.6, 
     font = 2,
     adj = c(0.60, 0))  

text(x = length(mean_val),
     y = mean_val[length(mean_val)]+0.025,
     labels = names(mean_val)[length(mean_val)],
     srt = 90, 
     cex = 0.6, 
     font = 2,
     adj = c(0, 0))  


```

## Extraction of award, honors, and teaching designations  

Comparison of MAJOR_SHORT and MAJOR, with some cleaning and manipulation, extracts the base degree awarded and whether or not an honors degree was awarded.  


```{r warning=FALSE}

# Difference between MAJOR and MAJOR_SHORT 

charDiff <- gsub(gradData$MAJOR_SHORT,  "", gradData$MAJOR)

charDiff <- mapply(function(x,y) {gsub(x, "", y) }, x = gradData$MAJOR_SHORT, y = gradData$MAJOR)

charLength <- sapply(charDiff, nchar)

charLength |> 
  table() |> 
  log() |>
  plot(ylab = "Count (log)",
       xlab = "Count of difference in characters",
       main = "Difference in character length \nbetween MAJOR_SHORT and MAJOR"
         )

eightY <- table(charLength) |> (\(x){x[names(x) == "8"]})() |> log()
text(x = 7,
      y = eightY+0.5,
      label = "B.A., B.S., etc",
     # pos= 3,
     adj = c(0,0)
     )


tenY <-  table(charLength) |> (\(x){x[names(x) == "10"]})() |> log()
text(x = 10,
      y = tenY + 0.3,
      label = "H.B.Mus., H.B.F.A., etc",
     # pos= 3,
     adj = c(0,0)
     )


sixteenY <- table(charLength) |> (\(x){x[names(x) == "16"]})() |> log()
text(x = 16,
      y = sixteenY + 0.3,
      label = "Psychology Teaching",
     # pos= 3,
     adj = c(0,0)
     )


thirtyFourY <- table(charLength) |> (\(x){x[names(x) == "34"]})() |> log()
text(x = 34,
      y = thirtyFourY + 1,
      label = "Needs cleaning",
     # pos= 3,
     adj = c(0.5,0.5)
     )


diff_exceptions_filter <- charLength > 20 

# table(gradData[diff_exceptions_filter, c("MAJOR_SHORT", "MAJOR")]) |> as.data.frame.matrix() |> View()

cbind(unique(gradData$MAJOR_SHORT[diff_exceptions_filter]),
      c("Consumer & Communities Studies",
        "Parks, Recreation, & Tourism",
        "Health, Society, & Policy",
        "Health Promotion & Education"
      )
) |>
  kableExtra::kbl(caption = "Mis-matched majors") 

```


```{r}  

# Extract Award

awards <- cleanDiff(majr = gradData$MAJOR, majr_short = gradData$MAJOR_SHORT) |>
 (\(x){ 
  gsub("[^[:alnum:]]", "", x)
})()  


honors <- ifelse(awards %in% c(
  "HBA",
  "HBS",
  "HBFA",
  "HBU",
  "HBMus",
  "HBSW"
),
"honors",
"regular"
)

base_awards <- gsub("H", "", awards) |>
  (\(x){
    gsub("^BU$","BUS",x)
  })()

# let's put these back on

gradData$award <- base_awards
gradData$honors <- honors

table(gradData[,c("award","honors")]) |>
  kableExtra::kbl(caption = "Basic and Honors degrees")




```

Another variation in awards is the "Teaching" designation.  It was awarded `r grepl("teach", tolower(gradData$MAJOR_LONG)) |> table() |> (\(x){x[names(x) == "TRUE"]})() |> format(big.mark = ",")` times (`r grepl("teach", tolower(gradData$MAJOR_LONG)) |> table() |> proportions() |> (\(x){x[names(x) == "TRUE"]})() |> (\(x){round(100*x,2) })()  |> format(big.mark = ",")`%).

```{r}

## TEACHING AWARDS ##

gradData$teach <- grepl("teach", tolower(gradData$MAJOR_LONG))

# unique(gradData$MAJOR_LONG[gradData$teach])

table(gradData[,c("award","teach")]) |>
  kbl(caption = "Teaching awards", col.names = c("", "Non-Teaching","Teaching"))


table(gradData[gradData$teach,c("MAJOR_SHORT")]) |>
    (\(x){x[order(x, decreasing = TRUE)]})() |>
  kbl(caption = "Teaching by major")

```

## Low Variance Columns Discarded

```{r}

## COLUMN VARIANCE ## 

library(skimr)

sapply(gradData, class) |> unlist() |> table()

# character   logical   numeric   POSIXct    POSIXt 
#       47         1        12         3         3 


charSkim <- skim(gradData[, sapply(gradData, is.character)])

# zero-variance character columns
# No need to query these
zv_col <- charSkim[charSkim$skim_type %in% c("character") & charSkim$character.n_unique == 1, 
  ] |>
  (\(x){x[,"skim_variable"][[1]]})() 

# > zv_col
# [1] "DEGREECAREER"            
# [2] "ACAD_DEPARTMENT_REG_SUPP"
# [3] "ACAD_DIVISION_REG_SUPP"  
# [4] "IPEDLEVEL" 


# low variance character columns
# defined as >80% mono-value

value_proportions <- gradData[,sapply(gradData, is.character)] |> #select character columns
  (\(x){x[,-which(colnames(x) %in% zv_col )]})() |> # remove zero variance columns
  (\(x){lapply(x,table)})() |> # table per character 
    (\(x){lapply(x, function(x){as.data.frame(x, stringsAsFactors = FALSE)})})() |> # convert to data frame
  (\(x){lapply(x, function(x){x[order(x[,"Freq"],decreasing = TRUE),]})})()  |> # descending order
  (\(x){lapply(x, function(x){ data.frame(Var1 = x[,"Var1"], prop = proportions(x[,"Freq"])) })})() # proportions


lv_col <- value_proportions |>
  (\(x){lapply(x, function(x){max(x[,"prop"])})})()  |> 
  (\(x){lapply(x, function(x){x>0.8} )})() |>
  unlist() |>
  (\(x){x[x]})() |>
  names()
  
value_proportions[lv_col]

# the problem is that these aren't my final columns.
# so these "low variance" columns could be the signal I'm looking for.

# I need to find these columns in ftfData, and the course data  

# These are candidates to decline to query, but I'd need to see their distribution on the incoming data.  This is the "outcome" data, as in, these people graduated.  A division type of "program" vs "department" could signal longer to take to graduate from, or have a higher failure rate. 


# The two type columns should be pretty similar--

detSim(gradData$"ACAD_DEPARTMENT_TYPE",gradData$"ACAD_DIVISION_TYPE") # 96% similar

# I'll pick "DIVISION_TYPE" as it has the higher variance.

numSkim <- skim(gradData[, sapply(gradData, is.numeric)])

# View(numSkim)

cv <- function(x) sd(x, na.rm=TRUE) / abs(mean(x, na.rm=TRUE))

num_var <- sapply(gradData[, sapply(gradData, is.numeric)], cv) |>
  (\(x){x[order(x)]})() |>
  round(3)

# > num_var
#  MAJOR_MIN_UNITS     IPEDSRATEYEAR     IPEDSCOMPYEAR 
#            0.011             0.027             0.027 
# MINUNITSREQUIRED           UOFUGPA     DEGREEUOFUGPA 
#            0.058             0.119             0.122 
#              AGE      UOFUGPAUNITS ROLLUP_SORT_ORDER 
#            0.212             0.337             0.347 
# NBRTERMSATTENDED            ALT_ID           FA_PELL 
#            0.412             0.658             2.074 


dateSkim <- skim(gradData[,
sapply(gradData, inherits, "POSIXt") | sapply(gradData, inherits, "POSIXct") ])

# all(gradData$DEGREEDATE == gradData$DEGREEDATE1)
# looks like I duplicated DEGREEDATE a column in my query -- I'll double-check and clean that out.

#> length(colnames(gradData))
#[1] 63
#> length(unique(colnames(gradData)))
#[1] 62



other_col <- c(
  "IPEDLEVEL",
  "TRANSFERFROMINSTATE",
  "IPEDSCOMPYEAR", 
  "IPEDSRATEYEAR", 
  "ALT_ID"
)



```

Zero-variance categorical columns are identified as `r zv_col`.  These have a single value and will not be queried.

Low-variance categorical columns are identified as `r lv_col`.  Of these, "ACAD_DEPARTMENT_TYPE" and "ACAD_DIVISION_TYPE" are 96.8% similar, and I will prefer "ACAD_DIVISION_TYPE" over  "ACAD_DEPARTMENT_TYPE" as it has higher variation.  

Low-variance numerical columns include MAJOR_MIN_UNITS and MINUNITSREQUIRED, both overwhelmingly at 122 units.  These both have a cross variance less than 0.1, and will not be selected. 


## Other Discarded Columns  

These columns are also discarded:  

  ALT_ID as it is an identifier and not descriptive 
  ROLLUP_SORT_ORDER as it appears to be related to data management and is not descriptive. 
  


# Results Detail 

### Reference

Some definitions are found in this [definitions document](https://registrar.utah.edu/_resources/documents/pdf/student-data-warehouse-definitions.pdf).

### Division

```{r}

## DIVISION ## 

lapply(simList[["div"]], function(x) { x[["sim"]]}) |> unlist() |> (\(x){
  
  idents <-  x[x=="Identical"]
  numerics <- x[x!="Identical"] |>
    as.numeric() |>
    round(3)
  names(numerics) <- names(x[x!="Identical"])
  
  numerics <- numerics[order(numerics, decreasing = TRUE)]
  
  c(
  idents,
  numerics
  )
  
  })()


```

### College

```{r}

## COLLEGE ## 

lapply(simList[["college"]], function(x) { x[["sim"]]}) |> unlist() |> (\(x){
  
  idents <-  x[x=="Identical"]
  numerics <- x[x!="Identical"] |>
    as.numeric() |>
    round(3)
  names(numerics) <- names(x[x!="Identical"])
  
  numerics <- numerics[order(numerics, decreasing = TRUE)]
  
  c(
  idents,
  numerics
  )
  
  })()

```

### Department

```{r}

## DEPT ## 

lapply(simList[["dept"]], function(x) { x[["sim"]]}) |> unlist() |> (\(x){
  
  idents <-  x[x=="Identical"]
  numerics <- x[x!="Identical"] |>
    as.numeric() |>
    round(3)
  names(numerics) <- names(x[x!="Identical"])
  
  numerics <- numerics[order(numerics, decreasing = TRUE)]
  
  c(
  idents,
  numerics
  )
  
  })()

```

### Major

```{r}

lapply(simList[["major"]], function(x) { x[["sim"]]})  |> unlist() |> (\(x){
  
  idents <-  x[x=="Identical"]
  numerics <- x[x!="Identical"] |>
    as.numeric() |>
    round(3)
  names(numerics) <- names(x[x!="Identical"])
  
  numerics <- numerics[order(numerics, decreasing = TRUE)]
  
  c(
  idents,
  numerics
  )
  
  })()


```


```{r eval=FALSE, include = FALSE}

# aborted attempt at cleaning

# unique(gradData$MAJOR[diff_exceptions_filter])

# View(gradData[diff_exceptions_filter,c("MAJOR_SHORT", "MAJOR")])

# Ok, I see the differences:
# 'Community" vs "Communities" 
# 'Recreation' vs "Recreation,'

table(gradData[diff_exceptions_filter,c("MAJOR_SHORT")])

#         Consumer & Community Studies 
#                                  506 
# Health Promotion & Education Teaching 
#                                   31 
#             Health, Society & Policy 
#                                 1145 
#          Parks, Recreation & Tourism 
#                                 1712 


table(gradData[diff_exceptions_filter,c("MAJOR")])

#  Consumer & Communities Studies, B.A. 
#                                    39 
#  Consumer & Communities Studies, B.S. 
#                                   465 
# Consumer & Communities Studies, H.B.S. 
#                                     2 
#  Health Promotion & Education, B.S.T. 
#                                    31 
#       Health, Society, & Policy, B.A. 
#                                    36 
#       Health, Society, & Policy, B.S. 
#                                  1036 
#     Health, Society, & Policy, H.B.A. 
#                                     5 
#     Health, Society, & Policy, H.B.S. 
#                                    68 
#    Parks, Recreation, & Tourism, B.A. 
#                                   115 
#    Parks, Recreation, & Tourism, B.S. 
#                                  1591 
#  Parks, Recreation, & Tourism, H.B.A. 
#                                     1 
#  Parks, Recreation, & Tourism, H.B.S. 
#                                     5 

charStripped <- charDiff |>
(\(x){ 
  gsub("[^[:alnum:]]", "", x)
})()

# Next I'll use the stripped characters to pull out the awards

busFilter <- charStripped == "BUS"
View(gradData[busFilter,c("MAJOR_SHORT","MAJOR")])

charStripped_length <- sapply(charStripped, nchar)

charStripped[charStripped_length <= 10] |> table()

charStripped[charStripped_length > 10] |> table()

# huh, looks like BST didn't make the cut

theAwards <- charStripped[charStripped_length <= 10] |> unique()


problemAwards <- charStripped[charStripped_length > 10] |> unique()

grepl(paste(theAwards, collapse = "|"), problemAwards)


```


```{r eval=FALSE, include=FALSE}


# testDiff <- cleanDiff(majr = gradData$MAJOR, majr_short = gradData$MAJOR_SHORT)

# testLength <- sapply(testDiff, nchar)

# fixed!


```


```{r include=FALSE, eval=FALSE}

simList

``` 

