---
title: "Column Selection from COMBINED_COURSE_V"
format: 
  html: default
  gfm: default
  pdf: default
  docx: default 
editor: source
execute: 
  warning: false  
knitr: 
  opts_chunk: 
    fig.height: 5
    fig.width: 7
---

**PURPOSE:**  The purpose of this report is to examine and select columns from COMBINED_COURSE_V table for further use and modeling.  It examines columns for similarity and variance, and compares with COMBINED_DEGREE_V.


```{r}

##########  
## LOAD ##  
########## 

courseData <- readRDS(here::here("Data", "courseData_8Apr26_graduated_ftf.rds"))

termExtract_Filter <- courseData$TERMEXTRACT %in% c("E","S")
courseData <- courseData[termExtract_Filter,] # apply filter

```


```{r}

###############
## FUNCTIONS ##
###############

library(kableExtra)
library(skimr)

detSim <- function(col1,col2) {

  # This function "determines similarity" between columns to determine if the columns are redundant or in other words if they are different labels for the same thing.  For example, "COLLEGE" and "COLLEGE_SHORT" should be completely redundant.  However, "COLLEGE" and "ACAD_COLLEGE_CIP_CD" might not be redundant.  
    
  testTable <- table(col1,col2) |>
    as.data.frame.matrix()
  
  # Every row and every column has exactly one non-zero cell
  identicalTest <- 
    all(apply(testTable, 1, function(x) sum(x > 0) == 1)) &&
    all(apply(testTable, 2, function(x) sum(x > 0) == 1))
  
  if(identicalTest) {return(
    list(sim = 1.000, row = NA, col = NA))
    }
  
  # return row and columns that are off 
  
  offRow <- apply(testTable, 1, function(x) sum(x > 0)) |> (\(x){which(x>1)})()
  offCol <- apply(testTable, 2, function(x) sum(x > 0)) |> (\(x){which(x>1)})()
  
  # percentage difference
  row_max_sum <- sum(apply(testTable, 1, max))
  col_max_sum <- sum(apply(testTable, 2, max))
  row_sim <- row_max_sum / sum(testTable)
  col_sim <- col_max_sum / sum(testTable)
  
  return(list(sim = min(row_sim, col_sim), row = offRow, col = offCol))
}


createCombinatorics <- function(vec) {

  # This function receives a vector of column names and then produces a grid of every dual combination.  It will be used to calculate the similarity
    
  full_grid <- expand.grid(vec,vec,stringsAsFactors = FALSE)
  reduced_grid <- full_grid[full_grid$Var1 < full_grid$Var2,]
  
  return(reduced_grid)
  
}


family_detSim <- function(fam, data){
  
  # where fam is a vector of column names chosen for their suspected similarity
  # where data is the dataframe where the columns occur
  
sim_list <- fam |>
createCombinatorics() |>
  (\(comb){
    setNames(
  lapply(1:nrow(comb),
         function(y){
           detSim(data[, comb[y,"Var1"]], data[,comb[y,"Var2"]])
         }
         ),
  paste(comb[,"Var1"],comb[,"Var2"])
    )
  })()

  return(sim_list)

}

## SIMILARITY SUMMARY ## 

sumSim <- function(y){
  
  # where "y" is output of family_detSim
  
  lapply(y, function(x) { x[["sim"]]} ) |> 
    unlist() |>
  (\(x){
  
  idents <-  x[x=="Identical"]
  numerics <- x[x!="Identical"] |>
    as.numeric() |>
    round(3)
  names(numerics) <- names(x[x!="Identical"])
  
  numerics <- numerics[order(numerics, decreasing = TRUE)]
  
  c(
  idents,
  numerics
  )
  
  })()

}  
  
meanSim <- function(sumSim_output, family = NA){
  
  if(any(is.na(family))){ # extract unique values
    
    family <- names(sumSim_output) |>
      strsplit(" ") |> 
      unlist() |>
      unique()
    
    }
  
  
mean_val <- lapply(family, function(fam_val){
 
  sumSim_output[grepl(fam_val, names(sumSim_output))] |>
    mean(na.rm = TRUE)
   
})
names(mean_val) <- family  

mean_val <- mean_val |>
  unlist() |>
  (\(x){x[order(x, decreasing = TRUE)]})()

return(mean_val) 
 
}


```

```{r}

############
## SAMPLE ##
############

# take a random sample (by EMPLID) to improve process time.

uniqueEmplid <- unique(courseData$EMPLID)

sample_list <- lapply(c(0.05,0.1,0.33,0.5), function(x) {
  
  sample(uniqueEmplid, x*length(uniqueEmplid))
  
})
names(sample_list) <- c("sample5", "sample10", "sample33", "sample50")

index_list <- lapply(sample_list, function(x){
  
  courseData$EMPLID %in% x
  
  
})


```


```{r}

#############
## COLUMNS ##
#############

# > colnames(courseData)
keepColumns <- c(
  "TERM",                        
  # [2] "ACADYR" # redundant to term                      
  "TERMEXTRACT",                 
  "MAX_SNAP",    # zero variance                
  "EOTDATE",                     
  # [6] "CENSUSDATE"  # not relevant; prefer EOTDATE              
  "EMPLID",                      
  # [8] "ALT_ID"        # not relevant                  
  # [9] "FULLNAME"      # redundant                
  "STUDENTCAREER",                
  "SUBJECT_CD",                  
  "SUBJECT_NAME",                
  "SUBJECT_LONG",                
  "SUBJECT_ACAD_ORG_CD",         
  "CATNBR",                      
  "CATNBR2",                     
  "SECTION",                     
  "TITLE",                       
  "UNITS",                       
  "GRADE",                       
  "INSTEMPLID",                  
  "INSTNAME",                    
# [23] "CLASSNBR"    # too granular                  
 "OFFERINGNBR",                 
 "SESSIONCODE",                  
 "COURSECAREER",   # "STUDENTCAREER" ?             
 "SCHEDFLAG",                   
 "WAUTOENRL",                   
 "AUTOENROLL",                  
 "COMPONENT",                   
 "GENED",                       
 "VARCREDIT",                   
 "CONTRACT",                    
 "DIRECTPAY",                   
 "CORRESPONDENCE",              
 "ONLINECOURSE",                
 "IVC",                         
 "IVC_HYBRID",                  
 "COURSE_MODALITY",             
 "INSTRUCTION_MODE",            
 "TELECOURSE",                  
 "STUDYABROAD",                 
 "EDNET",                       
 "HYBRIDCOURSE",                
 "COURSE_LEVEL",                
 "USHE_COURSE_LEVEL",           
# [47] "FISCAL_YEAR_OF_STARTDT" # redundant to TERM     
  "VP",                          
  "VP_SHORT",                    
  "VP_FORMAL",                   
  "ACAD_COLLEGE_CD",             
  "COLLEGE",                     
  "COLLEGE_SHORT",               
  "COLLEGE_FORMAL",              
  "ACAD_COLLEGE_REG_SUPP",       
  "ACAD_COLLEGE_TYPE",           
  "ACAD_COLLEGE_CIP_CD",         
  "ACAD_DEPARTMENT_CD",          
  "DEPARTMENT",                  
  "DEPARTMENT_SHORT",            
  "DEPARTMENT_FORMAL",           
  "ACAD_DEPARTMENT_REG_SUPP",    
  "ACAD_DEPARTMENT_TYPE",        
  "ACAD_DEPARTMENT_CIP_CD",      
  "ACAD_DIVISION_CD",            
  "DIVISION",                    
  "DIVISION_SHORT",              
  "DIVISION_FORMAL",             
  "ACAD_DIVISION_REG_SUPP",      
  "ACAD_DIVISION_TYPE",          
  "ACAD_DIVISION_CIP_CD",        
  "VP_CD",                       
  "COLLEGE_CD",                  
  "DEPARTMENT_CD",               
  "DIVISION_CD",                 
# [76] "ROLLUP_SORT_ORDER" # not relevant           
  "PS_ACAD_ORG",                
  "PS_ACAD_GROUP",               
  "CAMPUS",                      
  "COURSE_CAMPUS",               
  "USHE_SITE_TYPE_CD",           
  "USHE_SITE_TYPE",              
  "ONOFFCAMPUS",                 
  "COURSELOCATION",              
  "CONTACTMINUTES",              
  "TEAMTAUGHT",                  
  "XLIST",                       
  "BEGTIME1",                    
  "BEGTIME2",                    
  "BEGTIME3",                    
  "DAYS1",                       
  "DAYS2",                       
  "DAYS3",                       
  "ENDTIME1",                    
  "ENDTIME2",                    
  "ENDTIME3",                    
  "CLASSLOC1",                   
  "CLASSLOC2",                   
  "CLASSLOC3",                   
 "CLASSLOCBUILDNAME1",          
 "CLASSLOCBUILDROOM1",          
 "CLASSLOCBUILDNAME2",          
 "CLASSLOCBUILDROOM2",          
 "CLASSLOCBUILDNAME3",          
 "CLASSLOCBUILDROOM3",          
# [106] "STARTDT" # redundant to term                    
# [107] "ENDDT"   # redundant to term                    
 "BUDGETCODE",                  
 "LINEITEM" ,                   
 "SERVICELEARNING",             
 "GRADEGPA",                    
# [112] "XLIST_ID"  # prefer EMPLID                  
# [113] "COMBINEDID"  # prefer EMPLID                
# [114] "USHE_ACADYR" # redundant to term                
# [115] "USHE_TERM"   # redundant to term                
# [116] "TERM2"       # redundant to term                
#  [117] "ORG_EFFDT" # # redundant to term                   
 "CLASSENROLLMENTCAPACITY",     
 "ROOM_MAX_1",                  
 "ROOM_MAX_2",                  
 "ROOM_MAX_3",                  
 "TERM_NBR",                    
 "CLASS_ATTR_LIST",             
 "EXCLUDE_BUDGET_SCH",          
# [125] "SPR_CORRECTION_NOT_USHE_FLAG" # sounds irrelevant
# [126] "Section Divider: OLD"        # sounds irrelevant
 "SUBJECTCOLL",                 
 "SUBJECT"
)

```

```{r}

#####################
## COLUMN FAMILIES ##
#####################

# columns that sound like they are similar

subjectFamily <- c(
  "SUBJECT_CD",                  
  "SUBJECT_NAME",                
  "SUBJECT_LONG",                
  "SUBJECT_ACAD_ORG_CD",
  "SUBJECTCOLL",                 
  "SUBJECT"
  )

catFamily <- c(
  "CATNBR",
  "CATNBR2"
)

sectionFamily <- c(
  "SECTION",
  "OFFERINGNBR",                 
  "SESSIONCODE" 
)

careerFamily <- c(
  "STUDENTCAREER",
  "COURSECAREER"
)

enrollFamily <- c(
 "WAUTOENRL",                   
 "AUTOENROLL"  
)

levelFamily <-c(
 "COURSE_LEVEL",                
 "USHE_COURSE_LEVEL" 
)

vpFamily <- c(
 "VP",                          
 "VP_SHORT",                    
 "VP_FORMAL",
 "VP_CD" 
)

collegeFamily <- c(
  "ACAD_COLLEGE_CD",             
  "COLLEGE",                     
  "COLLEGE_SHORT",               
  "COLLEGE_FORMAL",              
  "ACAD_COLLEGE_REG_SUPP",       
  "ACAD_COLLEGE_TYPE",           
  "ACAD_COLLEGE_CIP_CD",
  "COLLEGE_CD",
  "SUBJECTCOLL",                 
  "SUBJECT"
)

deptFamily <- c(
 "ACAD_DEPARTMENT_CD",          
 "DEPARTMENT",                  
 "DEPARTMENT_SHORT",            
 "DEPARTMENT_FORMAL",           
 "ACAD_DEPARTMENT_REG_SUPP",    
 "ACAD_DEPARTMENT_TYPE",        
 "ACAD_DEPARTMENT_CIP_CD",
 "DEPARTMENT_CD"
)

divFamily <- c(
  "ACAD_DIVISION_CD",            
  "DIVISION",                    
  "DIVISION_SHORT",              
  "DIVISION_FORMAL",             
  "ACAD_DIVISION_REG_SUPP",      
  "ACAD_DIVISION_TYPE",          
  "ACAD_DIVISION_CIP_CD",
  "DIVISION_CD"
)

psFamily <- c(
  "PS_ACAD_ORG",                 
  "PS_ACAD_GROUP"
)

siteFamily <- c(
   "CAMPUS",                      
  "COURSE_CAMPUS",               
  "USHE_SITE_TYPE_CD",           
  "USHE_SITE_TYPE",              
  "ONOFFCAMPUS",                 
  "COURSELOCATION"              
)

begFamily <- c(
  "BEGTIME1",                    
  "BEGTIME2",                    
  "BEGTIME3"
)

dayFamily <- c(
 "DAYS1",                       
 "DAYS2",                       
 "DAYS3"
)

endFamily <- c(
 "ENDTIME1",                    
 "ENDTIME2",                    
 "ENDTIME3"
)

classLocFamily <- c(
  "CLASSLOC1",                   
  "CLASSLOC2",                   
  "CLASSLOC3",                   
  "CLASSLOCBUILDNAME1",          
  "CLASSLOCBUILDROOM1",          
  "CLASSLOCBUILDNAME2",          
  "CLASSLOCBUILDROOM2",          
  "CLASSLOCBUILDNAME3",          
  "CLASSLOCBUILDROOM3"        
)

roomMaxFamily <- c(
 "ROOM_MAX_1",                  
 "ROOM_MAX_2",                  
 "ROOM_MAX_3"
)

```

```{r}

## CALCULATE SIMILARITY ##

startTime <- Sys.time()
simList50 <- lapply(list(
  subject = subjectFamily,
  cat = catFamily,
  section = sectionFamily,
  career = careerFamily,
  enroll = enrollFamily,
  level = levelFamily,
  vp = vpFamily,
  
  div=divFamily,
  college =collegeFamily,
  dept = deptFamily,
  
  ps = psFamily,
  site = siteFamily,
  beg = begFamily,
  day = dayFamily,
  
  end = endFamily,
  classLoc = classLocFamily,
  room = roomMaxFamily
  
            ),
       family_detSim, data = courseData[index_list[["sample50"]],])
endTime <- Sys.time()

print(endTime - startTime)

```

```{r}

simList50_summary <- lapply(simList50, function(x){
  
  sumSim(x)
  
})

simList50_mean <- lapply(simList50_summary, function(x){
  
  meanSim(x)
  
})

```

# Results

```{r}

## REDUNDANCY RESULTS ## 

redundancy_results_list <- list(
  subject = data.frame(results = simList50_mean[["subject"]], keep = c(F,F,T,F,F,T)),
  cat = data.frame(results = simList50_mean[["cat"]], keep = c(T,F)),
  section = data.frame(results = simList50_mean[["section"]], keep = c(T,F,T)),
  career = data.frame(results = simList50_mean[["career"]], keep = c(F,T)),
  enroll = data.frame(results = simList50_mean[["enroll"]], keep = c(T,F)),
  level = data.frame(results = simList50_mean[["level"]], keep = c(T,T)),
  vp = data.frame(results = simList50_mean[["vp"]], keep = c(F,F,F,T)),
  
  div=data.frame(results = simList50_mean[["div"]], keep = c(T,F,F,F,F,T,T,T)),
  college =data.frame(results = simList50_mean[["college"]], keep = c(T,F,F,F,F,F,F,T,T,T)),
  dept = data.frame(results = simList50_mean[["dept"]], keep = c(T,F,F,F,F,F,T,T)),
  
  ps = data.frame(results = simList50_mean[["ps"]], keep = c(T,F)),
  site = data.frame(results = simList50_mean[["site"]], keep = c(T,F,F,F,F,F)),
  beg = data.frame(results = simList50_mean[["beg"]], keep = c(T,T,T)),
  day = data.frame(results = simList50_mean[["day"]], keep = c(T,T,T)),
  end = data.frame(results = simList50_mean[["end"]], keep = c(T,T,T)),
  classLoc = data.frame(results = simList50_mean[["classLoc"]], keep = c(T,T,T,T,T,T,T,T,T)),
  room = data.frame(results = simList50_mean[["room"]], keep = c(T,T,T)) 
)

redundancy_results_frame <- do.call(rbind, redundancy_results_list)
redundancy_results_frame$var <- sub("^[^.]*\\.", "",   rownames(redundancy_results_frame))

# rownames(result) <- gsub("^.*\\.", "", rownames(result))

# Now I'll use that, and the hand-selected columns removed,
# and then do the variance test on the remaining columns

redundantColumns <- redundancy_results_frame$var[!redundancy_results_frame$keep]

```


# Detecting zero, low variance, and high missing value columns

```{r}

# remove columns filtered out via redundancy tests 

testColumns <- keepColumns[!keepColumns %in% redundantColumns]

# sapply(courseData[,testColumns], class) |> unlist() |> table()

# character   numeric   POSIXct    POSIXt 
#       70         8         1         1 

## CHARACTER ##

charCols <- testColumns[sapply(courseData[,testColumns], is.character)]

charSkim <- skim(courseData[,charCols])

# zero variance categorical columns
zv_col_char <- charSkim[charSkim$skim_type %in% c("character") & charSkim$character.n_unique == 1, 
  ] |>
  (\(x){x[,"skim_variable"][[1]]})() 

# > zv_col_char
# [1] "MAX_SNAP"           "EXCLUDE_BUDGET_SCH"


# Empty columns

colCompletion_char <- courseData[,charCols] |>
  (\(x){x[,-which(colnames(x) %in% zv_col_char )]})() |> # remove zero variance columns
  (\(x){lapply(x, function(y) {sum(is.na(y))})} )() |>
     unlist() |>
  (\(x){(nrow(courseData) - x)/nrow(courseData)})() |>
  (\(x){x[order(x)]})()

# missing more than 75% of the values
empty_col_char <- colCompletion_char[colCompletion_char < 0.25] |>
  names()
   
# low variance categorical columns

value_proportions_char <- courseData[,charCols] |> #select character columns
  (\(x){x[,-which(colnames(x) %in% zv_col_char )]})() |> # remove zero variance columns
  (\(x){x[,-which(colnames(x) %in% empty_col_char )]})() |> # remove empty columns
  (\(x){lapply(x,table, useNA = "always")})() |> # table per character 
    (\(x){lapply(x, function(x){as.data.frame(x, stringsAsFactors = FALSE)})})() |> # convert to data frame
  (\(x){lapply(x, function(x){x[order(x[,"Freq"],decreasing = TRUE),]})})()  |> # descending order
  (\(x){lapply(x, function(x){ data.frame(Var1 = x[,"Var1"], prop = proportions(x[,"Freq"])) })})() # proportions

# where largest value is >90% of the column, including NA values
lv_col_char <- value_proportions_char |>
  (\(x){lapply(x, function(x){max(x[,"prop"])})})()  |> 
  (\(x){lapply(x, function(x){x>0.90} )})() |>
  unlist() |>
  (\(x){x[x]})() |>
  names()
  
# value_proportions[lv_col_char] |> length() # 19 # wow!

# What I think I'll do is ---

# Set a high rejection bar
# Train a model using these rejection columns
# Recover any useful columns

## NUMERIC ##

numCols <- testColumns[sapply(courseData[,testColumns], is.numeric)]

numSkim <- skim(courseData[,numCols])


# View(numSkim)

cv <- function(x) sd(x, na.rm=TRUE) / abs(mean(x, na.rm=TRUE))

num_var <- sapply(courseData[, numCols], cv) |>
  (\(x){x[order(x)]})() |>
  round(3)

#               GRADEGPA                TERM_NBR 
#                  0.257                   0.346 
#                  UNITS CLASSENROLLMENTCAPACITY 
#                  0.530                   1.089 
#         CONTACTMINUTES              ROOM_MAX_1 
#                  1.116                   1.128 
#             ROOM_MAX_2              ROOM_MAX_3 
#                  1.300                   1.304 

# None of these are rejected based on cross variance.

# Empty columns

colCompletion_num <- courseData[,numCols] |>
#  (\(x){x[,-which(colnames(x) %in% zv_col_char )]})() |> # no zero variance columns to remove
  (\(x){lapply(x, function(y) {sum(is.na(y))})} )() |>
     unlist() |>
  (\(x){(nrow(courseData) - x)/nrow(courseData)})() |>
  (\(x){x[order(x)]})()


# missing more than 75% of the values
empty_col_num <- colCompletion_num[colCompletion_num < 0.25] |>
  names()

#> empty_col_num
# [1] "ROOM_MAX_3" "ROOM_MAX_2" "ROOM_MAX_1"

# Date columns have already been removed, except for one that will be kept for manipulation

dateCols <- testColumns[sapply(courseData[, testColumns], function(x) {
  inherits(x, c("Date", "POSIXct", "POSIXlt", "POSIXt"))
})]

``` 

### Rejected columns

#### Redundant

```{r}

redundancy_results_frame$var[!redundancy_results_frame$keep]

```

#### Empty

```{r}

empty_col_char

empty_col_num

```

#### Zero variance

```{r}

zv_col_char

```

#### Low variance

```{r}

lv_col_char

```

### Selected columns

```{r}

finalKeep <- keepColumns[!keepColumns %in% c(redundantColumns, empty_col_char, empty_col_num, zv_col_char, lv_col_char)]

finalKeep


```

# Results Detail

## Redundancy

```{r}

redundancy_results_frame

# results_list

# simList50_summary

# simList50_mean

```

## Columns with completion rates less than 25%

```{r}

colCompletion_char[colCompletion_char < 0.25]


colCompletion_num[colCompletion_num < 0.25]



```

## Zero variance columns

```{r}

zv_col_char

```

## Low variance columns

```{r}

value_proportions_char[lv_col_char]

```



