---
title: "Query Validation Sandbox"
format: 
  html: default
  gfm: default
  pdf: default
  docx: default 
editor: source
execute: 
  echo: false
  warning: false  
  message: false
  error: false
knitr: 
  opts_chunk: 
    fig.height: 5
    fig.width: 7
---

**PURPOSE:**  The purpose of this document is to develop and investigate several ways to validate the query.  It is more akin to a "sandbox" rather than a formal report.

Ideas:

- I should look through my old work on course sequencing and DFWI's.   

- Compare my three unique terms queried to the grad/not_grad course data 
  - Try creating the same key in all of them and compare 
  - See if the EMPLID's are in the right place
  
- Try calculating the graduating GPA and see if it matches the grad data 

```{r}

##########
## LOAD ##
##########

## TERM FILES ##

term.files <- list.files(here::here("Data","Terms_14Apr26"))

termList <- lapply(term.files, function(x){
  
  readRDS(here::here("Data","Terms_14Apr26",x))
  
  })

termData <- do.call(rbind, termList)

## GRAD DATA ##

gradData <- readRDS(here::here("Data", "gradData_7Apr26.rds"))

## COURSE DATA ##

courseData_grad <- readRDS(here::here("Data", "courseData_8Apr26_graduated_ftf.rds"))

courseData_grad <- courseData_grad[courseData_grad$TERMEXTRACT %in% c("E","S"),]

courseData_not <- readRDS(here::here("Data", "courseData_8Apr26_NOT_graduated_ftf.rds"))

courseData_not <- courseData_not[courseData_not$TERMEXTRACT %in% c("E","S"),]

```

```{r}

###############
## FUNCTIONS ##
###############

findDupes <- function(data){
  
  
  # First, find duplicate keys
  theDupes <- table(data[,"key"], useNA = "always") |>
    (\(x){x[order(x, decreasing = TRUE)]})() |>
    (\(x){x[x>1]})() |>
    names()
  
}


```


```{r}

## SIMPLE EMPLID CHECK ##

# all EMPLID's are consistently located.  GOOD!

all(termData$EMPLID %in% gradData$EMPLID) # FALSE (as expected)

all(termData$EMPLID[as.logical(termData$IN_DEGREE_TABLE)] %in% gradData$EMPLID) # TRUE # good 

any(termData$EMPLID[!as.logical(termData$IN_DEGREE_TABLE)] %in% gradData$EMPLID) # FALSE # good

# ok, this is looking good so far.

all(termData$EMPLID %in% c(courseData_grad$EMPLID, courseData_not$EMPLID)) # TRUE # good

all(termData$EMPLID[as.logical(termData$IN_DEGREE_TABLE)] %in% courseData_grad$EMPLID ) # TRUE # good

any(termData$EMPLID[!as.logical(termData$IN_DEGREE_TABLE)] %in% courseData_grad$EMPLID ) # FALSE # good

any(termData$EMPLID[as.logical(termData$IN_DEGREE_TABLE)] %in% courseData_not$EMPLID) # FALSE # good

all(termData$EMPLID[!as.logical(termData$IN_DEGREE_TABLE)] %in% courseData_not$EMPLID) # TRUE # good


```


```{r}

## TERM DATA KEY ##

termData$key <- paste(termData$EMPLID,
                      termData$TERM,
                      termData$CATNBR,
                    #  termData$INSTEMPLID,
                    #  termData$COMPONENT,
                      termData$SECTION,
                    termData$SUBJECT_CD,
                      sep = "_")

length(unique(termData$key))/nrow(termData) 
# EMPLID only # 0,12 # should be multiple classes per EMPLID 
# add TERM    # 0,21 # should be multiple classes per EMPLID per term 
# add CATNBR  # 0.867 # maybe multiple instructors or multiple sections (like labs) 
# add INSTEMPLID # 0,908 # a small improvement
# add COMPONENT # 0.999 # A few stubborn exceptions of multiple identical labs (upon inspection one example is 'bowling')
# add SECTION # 1.0 # there it is!
# remove COMPONENT # 1.0 # Can drop 'COMPONENT'; turns out you can't duplicate sections per CATNBR per student
# remove INSTEMPLID # 0.994 # only a few sections with multiple instructors
# add SUBJECTCD # 1 # so no multiple teachers per code/catnbr/section ?

# NEXT QUESTION ---

# Should I remove the INSTEMPLID ?


termData_subset <- termData[,-which(colnames(termData) %in% c("INSTEMPLID")) ] |> unique()

termData_subset$key <- paste(termData_subset$EMPLID,
                      termData_subset$TERM,
                      termData_subset$CATNBR,
                      termData_subset$SUBJECT_CD,
                    #  termData$INSTEMPLID,
                    #  termData$COMPONENT,
                      termData_subset$SECTION,
                      sep = "_")

length(unique(termData_subset$key))/nrow(termData_subset) 


findDupes(termData_subset) |>
  (\(x){sample(x,12)})() |> # select a dozen keys
  (\(x){termData_subset[termData_subset$key %in% x,]})() |> # subset 
  (\(x){x[,c(ncol(x), 1:(ncol(x)-1)) ]})() |> # rearrange column order
  (\(x){x}[order(x$key), ])() |>
  View(title = "dupes")

findDupes(termData) |>
  (\(x){View(termData[termData$key %in% x,])})()


```

For simplicity, I think I'll have to drop the INSTEMPLID -- although I expect that the INSTEMPLID will affect completion rates, since they have grade variation after all.  

Or, rather than for simplicity, for the scope of the problem.  The "instructor-agnostic" philosophy probably won't change.

Looks like the COMPONENT == "LAB"  might be something I filter on, or UNITS != 0

```{r}

# checking termData$COMPONENT == "LAB"

# > sum(is.na(termData$COMPONENT))
# [1] 0

# > sum(is.na(termData$GRADEGPA))
# [1] 36053

mean(termData$GRADEGPA[termData$COMPONENT == "LAB"], na.rm=TRUE) # 3.67

# so I have NA values for GRADEGPA
# and I have grades for labs.

# Great, just great

mean(termData$UNITS[termData$COMPONENT == "LAB"], na.rm=TRUE) # 0.218

# is there any GPA when units are zero?

mean(termData$GRADEGPA[termData$UNITS == 0], na.rm=TRUE)  # NaN

sum(!is.na(termData$GRADEGPA[termData$UNITS == 0])) # 0

# Good, at least that check of internal consistency passed.

# Do I only want classes that have a UNIT value?  Maybe that's my filter?

table(termData$COMPONENT[termData$UNITS == 0])

#  ACT   CLN   CNT   DIS   ENS   FLD   FLW   IND   L/D   LAB   LEC 
#  282     1     3  7742     3     4    52     2    27 11596  1914 
#  NET   SEM   STU   TPC   WRK 
#    7     8    40     4    12 

table(termData$COMPONENT[termData$UNITS != 0])

#   ACT    CLN    DIS    ENS    FLD    FLW    HON    IND    L/D 
#  4945   1200    513   1109     17   1223    211    686     25 
#   LAB    LEC    PRA    PRJ    PRV    RSC    SEM    STU    THE 
#  2309 115469   1781    672    648     94   9964   4781    328 
#   TPC    WRK 
#  3923    375 

# seems like adding a WHERE UNITS != 0 might be reasonable for a query, 
# or I might pass a column with a flag like "HAS_LAB" 

```



```{r}

## SECTION ##

# If I am adding "SECTION" to my key,
# seems like there's a class number I should just bring in!



```


```{r}

## COURSE DATA KEY (grad) ## 

courseData_grad_subset <- courseData_grad[courseData_grad$TERM %in% unique(termData$TERM),]

courseData_grad_subset$key <- paste(
                    courseData_grad_subset$EMPLID,
                      courseData_grad_subset$TERM,
                      courseData_grad_subset$SUBJECT_CD,
                      courseData_grad_subset$CATNBR,
                    #  courseData_grad_subset$INSTEMPLID,
                    #  termData$COMPONENT,
                      courseData_grad_subset$SECTION,
                      sep = "_")

length(unique(courseData_grad_subset$key))/nrow(courseData_grad_subset)
# 1 # good key, I guess

dim(courseData_grad_subset)
# [1] 140299     129

theDupes <- findDupes(courseData_grad_subset)

length(theDupes) # 0 

findDupes(courseData_grad_subset) |>
  (\(x){sample(x,12)})() |> # select a dozen keys
  (\(x){courseData_grad_subset[courseData_grad_subset$key %in% x,]})() |> # subset 
  (\(x){x[,c(ncol(x), 1:(ncol(x)-1)) ]})() |> # rearrange column order
  (\(x){x}[order(x$key), ])() |>
  View(title = "dupes")

courseData_grad_subset$key2 <- paste(
                    courseData_grad_subset$EMPLID,
                      courseData_grad_subset$TERM,
                      courseData_grad_subset$CATNBR,
                      courseData_grad_subset$INSTEMPLID,
                    #  termData$COMPONENT,
                      courseData_grad_subset$SECTION,
                      sep = "_")


```
